---
title: "Mapping Linguistic Diversity"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Mapping Linguistic Diversity}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 8,
  fig.height = 10,
  warning = FALSE,
  message = FALSE
)
```

India is one of the most linguistically diverse countries in the world. The 2011 Census recorded over 19,500 languages and dialects. This vignette demonstrates how to visualize this diversity using Shannon entropy.

## Load data

```{r load-packages}
library(indiacensus)
library(dplyr)
```

```{r load-data}
data(census_2011_mother_tongue)

census_2011_mother_tongue |>
  filter(district_code == "000") |>
  group_by(language_level) |>
  summarise(n_languages = n_distinct(language_name))
```

The C-16 tables classify languages as:

- **L1**: 22 Scheduled Languages recognized in the Constitution
- **L2**: Dialects and variants within each scheduled language

## Calculate Shannon entropy

The **effective number of languages** ($2^H$) gives an intuitive interpretation: a district with $2^H = 4$ has diversity equivalent to four equally-spoken languages.

```{r calculate-entropy}
calculate_diversity <- function(data, level) {
  data |>
    filter(language_level == level, district_code != "000", !grepl("Others", language_name)) |>
    group_by(state_code, state_name, state_name_harmonized, district_code, area_name) |>
    mutate(proportion = total_persons / sum(total_persons)) |>
    summarise(
      n_languages = n(),
      shannon_entropy = -sum(proportion * log2(proportion), na.rm = TRUE),
      effective_languages = 2^shannon_entropy,
      .groups = "drop"
    )
}

diversity_l1 <- calculate_diversity(census_2011_mother_tongue, "L1")
diversity_l2 <- calculate_diversity(census_2011_mother_tongue, "L2")

cat("L1 (Scheduled Languages):\n")
summary(diversity_l1$effective_languages)

cat("\nL2 (Dialects):\n")
summary(diversity_l2$effective_languages)
```

## Prepare map data

```{r prepare-maps}
districts_sf <- get_census_boundaries(2011, "district") |>
  mutate(district_key = tolower(gsub("[^a-z0-9 ]", "", district_name)))

prepare_diversity <- function(div_data) {
  div_data |>
    mutate(district_key = tolower(gsub("[^a-z0-9 ]", "", gsub(" District$", "", area_name))))
}

map_l1 <- left_join(
  districts_sf,
  prepare_diversity(diversity_l1) |> select(state_name_harmonized, district_key, effective_languages),
  by = c("state_name_harmonized", "district_key")
)

map_l2 <- left_join(
  districts_sf,
  prepare_diversity(diversity_l2) |> select(state_name_harmonized, district_key, effective_languages),
  by = c("state_name_harmonized", "district_key")
)
```

## L1: Scheduled languages diversity

```{r plot-l1, fig.width=10, fig.height=12}
plot_map(
  map_l1,
  fill_var = "effective_languages",
  title = "Linguistic Diversity: Scheduled Languages (L1)",
  subtitle = "Effective number of languages per district",
  legend_title = "Effective\nLanguages",
  palette = "reds",
  show_state_boundaries = TRUE,
  trans = "log2",
  limits = c(1, 18),
  breaks = c(1, 2, 4, 8, 16)
)
```

## L2: Dialect-level diversity

```{r plot-l2, fig.width=10, fig.height=12}
plot_map(
  map_l2,
  fill_var = "effective_languages",
  title = "Linguistic Diversity: Dialects (L2)",
  subtitle = "Effective number of dialects per district",
  legend_title = "Effective\nLanguages",
  palette = "reds",
  show_state_boundaries = TRUE,
  trans = "log2",
  limits = c(1, 35),
  breaks = c(1, 2, 4, 8, 16, 32)
)
```

## Most diverse districts

```{r compare-levels}
cat("Top 10 by Scheduled Languages (L1):\n")
diversity_l1 |>
  arrange(desc(effective_languages)) |>
  select(state_name_harmonized, area_name, effective_languages) |>
  head(10)

cat("\nTop 10 by Dialects (L2):\n")
diversity_l2 |>
  arrange(desc(effective_languages)) |>
  select(state_name_harmonized, area_name, effective_languages) |>
  head(10)
```

## Regional patterns

**L1 diversity** is highest in cosmopolitan cities and Northeast states.

**L2 diversity** reveals hidden variation in the Hindi belt, where dialects like Bhojpuri, Maithili, and Marwari are classified under "Hindi" at L1.
